---
title: "Computational Techniques Assignment 8"
author: "Adeline Makokha - 191199"
date: "2025-05-12"
output: html_document
---

## Question 1: Hospital Department Efficiency

You are analyzing the performance of 4 hospital departments. Each department has a different combination of medical staff and beds.

**Objective:**  
Determine which department is most efficient in treating patients with limited resources.

**Inputs:**
- Doctors: 40, 35, 45, 38  
- Nurses: 80, 70, 85, 75  
- Beds: 150, 140, 160, 145  

**Outputs:**
- Patients treated: 2000, 1900, 2100, 1950  
- Survival rate: 0.97, 0.96, 0.98, 0.95  

---





```{r}
library(Benchmarking)

# Read hospital data

hospital_data <- read.csv("C:/Users/PC/Downloads/ADEH-MSC/Module 3/Module-3/DSA 8302 Computational Techniques in DS/week 8/hospital_data.csv")

```



```{r}
# Print raw data
print("Hospital Department Data:")
print(hospital_data)

# Prepare input and output matrices
inputs <- as.matrix(hospital_data[, c("Doctors", "Nurses", "Beds")])
outputs <- as.matrix(hospital_data[, c("Patients", "SurvivalRate")])

# DEA using CCR model (constant returns to scale), input-oriented
dea_ccr <- dea(X = inputs, Y = outputs, RTS = "crs", ORIENTATION = "in")
hospital_data$CCR_Efficiency <- dea_ccr$eff

# DEA using BCC model (variable returns to scale), input-oriented
dea_bcc <- dea(X = inputs, Y = outputs, RTS = "vrs", ORIENTATION = "in")
hospital_data$BCC_Efficiency <- dea_bcc$eff

# Print efficiency results
print("DEA Efficiency Scores (CCR and BCC):")
print(hospital_data[, c("DMU", "CCR_Efficiency", "BCC_Efficiency")])

# Plot CCR Efficiency
barplot(hospital_data$CCR_Efficiency, names.arg=hospital_data$DMU,
        col="steelblue", ylim=c(0,1.1), main="CCR Efficiency Scores",
        ylab="Efficiency Score", las=2)

# Plot BCC Efficiency
barplot(hospital_data$BCC_Efficiency, names.arg=hospital_data$DMU,
        col="darkgreen", ylim=c(0,1.1), main="BCC Efficiency Scores",
        ylab="Efficiency Score", las=2)

```

## Question 2: Retail Store Efficiency

A retail company wants to assess whether its 4 branches are efficiently turning staff and marketing investment into profit.

**Objective:**  
Identify the best performing store branch given similar resource levels.

**Inputs:**
- Marketing Spend (USD): 12,000; 10,000; 11,500; 9,800  
- Employees: 20, 18, 22, 19  

**Outputs:**
- Monthly Revenue (USD): 110,000; 100,000; 120,000; 95,000  
- Number of Customers: 2,300; 2,000; 2,600; 1,900  

---


```{r}
# Read retail data

retail_data <- read.csv("C:/Users/PC/Downloads/ADEH-MSC/Module 3/Module-3/DSA 8302 Computational Techniques in DS/week 8/retail_data.csv")

# Print raw data
print("Retail Store Data:")
print(retail_data)

# Prepare input and output matrices
inputs <- as.matrix(retail_data[, c("MarketingSpend", "Employees")])
outputs <- as.matrix(retail_data[, c("Revenue", "Customers")])

# DEA using CCR model (constant returns to scale), input-oriented
dea_ccr <- dea(X = inputs, Y = outputs, RTS = "crs", ORIENTATION = "in")
retail_data$CCR_Efficiency <- dea_ccr$eff

# DEA using BCC model (variable returns to scale), input-oriented
dea_bcc <- dea(X = inputs, Y = outputs, RTS = "vrs", ORIENTATION = "in")
retail_data$BCC_Efficiency <- dea_bcc$eff

# Print efficiency results
print("DEA Efficiency Scores (CCR and BCC):")
print(retail_data[, c("DMU", "CCR_Efficiency", "BCC_Efficiency")])

# Plot CCR Efficiency
barplot(retail_data$CCR_Efficiency, names.arg=retail_data$DMU,
        col="steelblue", ylim=c(0,1.1), main="CCR Efficiency Scores - Retail",
        ylab="Efficiency Score", las=2)

# Plot BCC Efficiency
barplot(retail_data$BCC_Efficiency, names.arg=retail_data$DMU,
        col="darkgreen", ylim=c(0,1.1), main="BCC Efficiency Scores - Retail",
        ylab="Efficiency Score", las=2)

```

## Question 3: School Performance Evaluation

A school district monitors 4 schools to determine which is using its resources most efficiently to produce better academic results.

**Inputs:**
- Teachers: 30, 25, 20, 28  
- Annual Budget (USD): 800,000; 750,000; 700,000; 770,000  

**Outputs:**
- Graduation Rate (%): 85, 90, 82, 88  
- Average Test Score (out of 100): 78, 80, 76, 79


```{r}
# Read school data

school_data <- read.csv("C:/Users/PC/Downloads/ADEH-MSC/Module 3/Module-3/DSA 8302 Computational Techniques in DS/week 8/school_data.csv")

# Print raw data
print("School Performance Data:")
print(school_data)

# Prepare input and output matrices
inputs <- as.matrix(school_data[, c("Teachers", "Budget")])
outputs <- as.matrix(school_data[, c("GraduationRate", "TestScore")])

# DEA using CCR model (constant returns to scale), input-oriented
dea_ccr <- dea(X = inputs, Y = outputs, RTS = "crs", ORIENTATION = "in")
school_data$CCR_Efficiency <- dea_ccr$eff

# DEA using BCC model (variable returns to scale), input-oriented
dea_bcc <- dea(X = inputs, Y = outputs, RTS = "vrs", ORIENTATION = "in")
school_data$BCC_Efficiency <- dea_bcc$eff

# Print efficiency results
print("DEA Efficiency Scores (CCR and BCC):")
print(school_data[, c("DMU", "CCR_Efficiency", "BCC_Efficiency")])

# Plot CCR Efficiency
barplot(school_data$CCR_Efficiency, names.arg=school_data$DMU,
        col="steelblue", ylim=c(0,1.1), main="CCR Efficiency Scores - Schools",
        ylab="Efficiency Score", las=2)

# Plot BCC Efficiency
barplot(school_data$BCC_Efficiency, names.arg=school_data$DMU,
        col="darkgreen", ylim=c(0,1.1), main="BCC Efficiency Scores - Schools",
        ylab="Efficiency Score", las=2)



```